---
title: "Introducing ggplot"
format:
  html:
    theme: cosmo
    highlight-style: monochrome
    title-block-banner: "#ffdc00"
    number-depth: 2
---

<!--  pdf:
    papersize: A4
    documentclass: memoir
    mainfont: Roboto
    sansfont: Clancy
    include-in-header: preamble.tex
    highlight-style: monochrome
---
-->


```{r}
#| message: false
#| warning: false
#| include: false

pbc <- readRDS("data/activities/mod01_pbc.rds")

pbc$sex <- factor(pbc$sex,
    levels = c(1, 2),
    labels = c("Male", "Female"))

pbc$stage <- factor(pbc$stage,
                    levels = c(1, 2, 3, 4),
                    labels = c("Stage 1", "Stage 2", "Stage 3", "Stage 4"))
```


## Introducing `ggplot2`

Plots produced by the `plot(dens())` and `boxplot()` functions have used so-called `base` graphics: no special packages have been needed to produce these plots. In this section, we will introduce the `ggplot2` [^1] package, which defines plots using a **grammar of graphics**. This can seem like a complex way of constructing a plot, but the framework allows for very complex plots to be constructed in a relatively consistent way.

The grammar of graphics defines a number of elements of a plot; the key elements that we will use in this course are:

- data: we need to specify the data we are plotting. The data must be stored as a dataframe, where variables are organised in columns, and observations are in rows
- mapping: the *aesthetic mapping* describes how your data map to visual properties of the plot. For example, we might want to define the x-axis by the `age` variable and the colour of points by `sex`. We can define our mapping using the command `mapping = aes()`, and entering the specific aesthetic components within the `aes()` component.
- geometries: we can think of *geometries* as defining the actual "ink" of the plot, modified by the *aesthetics* defined by the mapping. For example:
  - `geom_bar()` will plot a bar-chart;
  - `geom_boxplot()` will plot a boxplot;
  - `geom_histogram()` will plot a histogram.
- theme: the *theme* defines the overall look of the plot. Every part of the plot that is not specified by data can be adjusted by changing the theme. For example, fonts, axis ticks, legend position and background colours can all be changed by using a theme. Some useful themes are:
  - `theme_bw()`: a black-and-white theme
  - `theme_minimal()`: a clean them with no background annotations
  - `theme_classic()`: a theme that removes gridlines and backrounds, leaving only axis lines.

These elements can be combined in a single `ggplot` command by using a `+` symbol, usually with each element provided on a new line.

### Using `ggplot2` to construct a bar chart

As a demonstration, we will produce the simplest bar chart possible: a bar chart of `sex`.

First, we will load the `ggplot2` library, and call the `ggplot` function with no function arguments.

```{r}
library(ggplot2)

ggplot()
```

In this example, all `ggplot` has done is to create a blank canvas. Let's start filling the canvas in by specifying our dataset, and defining sex as our x-axis:

```{r}
ggplot(data=pbc, mapping = aes(x=sex))
```

Our blank canvas has been altered: we now have an x-axis, defined by `sex`, with values of `Male` and `Female`.

To add "ink" to our canvas, we now need to add a *geometry*. For a bar chart, we use `geom_bar()` as follows:

```{r}
ggplot(data=pbc, mapping = aes(x=sex)) +
  geom_bar()
```

We can change the look of the figure by choosing a different *theme*:let's use `theme_classic()` here.

```{r}
ggplot(data=pbc, mapping = aes(x=sex)) +
  geom_bar() +
  theme_classic()
```

The final changes we can make are to define the axis labels using `labs()`:

```{r}
ggplot(data=pbc, mapping = aes(x=sex)) +
  geom_bar() +
  theme_classic() +
  labs(x = "Participant sex", 
       y = "Frequency")
```

Finally, we can change the colour of the bars by specifying a `fill` colour within the `geom_bar()` command. We can choose one of over 600 (!) colours by name [^2].

```{r}
ggplot(data=pbc, mapping = aes(x=sex)) +
  geom_bar(fill="steelblue") +
  theme_classic() +
  labs(x = "Participant sex", 
       y = "Frequency")
```

Let's adapt this code to produce a bar chart for stage of disease:

```{r}
ggplot(data=pbc, mapping = aes(x=stage)) +
  geom_bar(fill="steelblue") +
  theme_classic() +
  labs(x = "Stage of disease", 
       y = "Frequency")
```

Note that `ggplot` has included a separate bar for the missing values (i.e. the `NA` values). We can instruct R to ignore the missing values in the `ggplot` function by specifying the data to be a `subset` of `pbc`. The command: `subset(pbc, is.na(stage) == FALSE)` essentially creates a temporary subset of the `pbc` dataset where `stage` is not missing. Putting it all together:

```{r}
ggplot(data=subset(pbc, is.na(stage) == FALSE), mapping = aes(x=stage)) +
  geom_bar(fill="steelblue") +
  theme_classic() +
  labs(x = "Stage of disease", 
       y = "Frequency")
```

## Introducing `|>`: aka the "pipe"

Consider the previous example, which essentially combines two functions: `ggplot()` and `subset()` by using: `ggplot(subset())` - in other words, `subset()` has been nested within `ggplot()`. When R comes across nested functions, it will happily run the inner-most function first, then applies the outer function to the result of the inner function.

While this code is understandable by R, one could argue this code is difficult to understand for a person. A more interpretable version of this code for a person would be:

- first run the `subset()` function, **and then**
- run the `ggplot()` function.

This can be rewritten using the `|>` symbol:

```
subset() |> 
  ggplot()
```

This syntax would "pipe" the results of the first function into the second function, and produces code that is much easier for a person to read. So our previous bar chart of stage of disease can be rewritten as:

```{r}
subset(pbc, is.na(stage) == FALSE) |> 
  ggplot(mapping = aes(x=stage)) +
    geom_bar(fill="steelblue") +
    theme_classic() +
  labs(x = "Stage of disease", 
       y = "Frequency")
```

Notice that when we pipe a dataset from the `subset()` function, we no longer need to specify the name of the dataset in `ggplot()`: the dataset is inferred from the action of the `subset()` function.

::: {.callout-note}
The `|>` pipe was introduced in R version 4.1.0, May 2021, and for this author (TD), has drastically improved the code that I write (and read!). When reading code that I, or others, have written using the pipe, I replace `|>` by the words **and then**. So the previous code would be read: subset the data **and then** plot the data.

You may also see `%>%` referred to as a pipe. This version of the pipe was introduced in the early 2010s, most notably in the `magrittr` package. While there are subtle differences between `|>` and `%>%`, the `|>` is the easier to use for this course, as it does not require loading any additional packages. More in-depth history can be found at [this blog post](http://adolfoalvarez.cl/blog/2021-09-16-plumbers-chains-and-famous-painters-the-history-of-the-pipe-operator-in-r/).
:::

### Using `ggplot2` to construct bar charts for two variables

In Section X.X, we saw hat clustered or stacked bar charts can be used to summarise two categorical variables. Remember that the *aesthetic mapping* can be used to describe how our data map to visual elements of our plot. So we can map `sex` to the `fill` attribute to summarise Stage of Disease by sex. By default, `ggplot` will produce a stacked bar chart, but we can request a clustered bar chart by specifying `position="dodge"` in the `geom_bar()` statement:

```{r}
subset(pbc, is.na(stage) == FALSE) |> 
  ggplot(mapping = aes(x=stage, fill=sex)) +
    geom_bar(position = "dodge") +
    theme_classic() +
  labs(x = "Stage of disease", 
       y = "Frequency")
```

Notice that the legend title uses the name of the `fill` variable, in this case `sex` with a lower-case s. We can specify an alternative legend title in the `labs()` statement:

```{r}
subset(pbc, is.na(stage) == FALSE) |> 
  ggplot(mapping = aes(x=stage, fill=sex)) +
    geom_bar(position = "dodge") +
    theme_classic() +
  labs(x = "Stage of disease", 
       y = "Frequency",
       fill = "Sex")
```

A stacked bar chart can be constructed by removing the `position = dodge()` statement:

```{r}
subset(pbc, is.na(stage) == FALSE) |> 
  ggplot(mapping = aes(x=stage, fill=sex)) +
    geom_bar() +
    theme_classic() +
  labs(x = "Stage of disease", 
       y = "Frequency",
       fill = "Sex")
```

Finally, we can construct a stacked relative bar chart by specifying `position = "fill"` in the `geom_bar()` definition:

```{r}
subset(pbc, is.na(stage) == FALSE) |> 
  ggplot(mapping = aes(x=stage, fill=sex)) +
    geom_bar(position = "fill") +
    theme_classic() +
  labs(x = "Stage of disease", 
       y = "Proportion",
       fill = "Sex")
```

## Producing a density plot using ggplot

We can produce a density plot by specifying the variable of interest as the `x` variable, and requesting a `geom_density()`:

```{r}
ggplot(data=pbc, aes(x=age)) + 
  geom_density() +
  theme_classic() +
  labs(x="Age (years)",
       y="Density")
```

## Producing a boxplot using ggplot

A boxplot is constructed with a single `y` variable, and `geom_boxplot()`:

```{r}
ggplot(data=pbc, aes(y=age)) + 
  geom_boxplot() +
  theme_classic() +
  labs(y="Age (years)")
```

By default, `ggplot2` includes arbitrary values on the x-axis. As boxplots are usually constructed only for informal data investigation, and are rarely included in formal documents, the x-axis labelling may not be an issue. However, the x-axis labelling can be removed as follows:

```{r}
ggplot(data=pbc, aes(y=age)) + 
  geom_boxplot() +
  theme_classic() +
  labs(y="Age (years)") +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```


[^1]: Note that the package is called `ggplot2` as it is a complete overhaul of the package called `ggplot`. The main function used in the `ggplot2` package is `ggplot()`. This can be confusing, but you will get used to this in time.
[^2]: A list of available colours that can be used in `ggplot` can be found by entering the command `grDevices::colours()` in R, or by searching the web for `R colour names`.