## Checking data for the independent samples t-test

### Examining variable distributions by a second variable

We can use the a `splitBy` variable in the `descriptives` function in the `jmv` package to obtain summary statistics for each level of a grouping variable. Further, specifying `dens = TRUE` will produce density plots for the analysis variable for each level of the grouping variable.

For example, to create the distribution plots in @fig-mod05-bwt-gender, we can use

```{r}
library(jmv)

bwt <- readRDS("data/examples/mod05_birthweight.rds")

descriptives(data = bwt, vars = birthweight, splitBy = gender, dens = TRUE)
```

Of course, we could also use `ggplot2` to produce the density plots, defining `gender` as a `facet_wrap`:

```{r}
library(ggplot2)

ggplot(data=bwt, aes(x=birthweight)) +
  geom_density() +
  facet_wrap(vars(gender)) +
  theme_classic() +
  labs(x="Birthweight (kg)")
```


## Independent samples t-test

We can use the `ttestIS()` (t-test, independent samples) function from the `jmv` package to perform the independent samples t-test. We include the `meanDiff=TRUE` and `ci=TRUE` options to obtain the difference in means, with its 95% confidence interval. We can request a Welch's test (which does not assume equal variances) by the `welchs=TRUE` option (note that output has been edited to fit on the page):

```
ttestIS(data = bwt, vars = birthweight, group = gender, 
        meanDiff = TRUE, ci = TRUE, welchs = TRUE)


INDEPENDENT SAMPLES T-TEST

 Independent Samples T-Test                                                                                                          
─────────────────────────────────────────────────────────────────────
                                 Statistic    df          p           
─────────────────────────────────────────────────────────────────────
   birthweight    Student's t     2.296556    98.00000    0.0237731   
                  Welch's t       2.303840    93.54377    0.0234458   
─────────────────────────────────────────────────────────────────────
────────────────────────────────────────────────────────────────────────────── 
                Mean difference    SE difference    Lower         Upper       
────────────────────────────────────────────────────────────────────────────── 
  Student's t        0.1660471       0.07230265    0.02256481    0.3095293   
  Welch's t          0.1660471       0.07207403    0.02293328    0.3091609   
────────────────────────────────────────────────────────────────────────────── 
   Note. Hₐ μ <sub>Female</sub> ≠ μ <sub>Male</sub>
```


## Checking the assumptions for a Paired t-test

Before performing a paired t-test, you must check that the assumptions for the test have been met. Using the dataset `mod05_ankle_bp.xlsx` to show that the difference between the pair of measurements between the sites is normally distributed, we first need to compute a new variable of the differences and examine its distribution.

```{r}
library(readxl)
library(dplyr)

sbp <- read_excel("data/examples/mod05_ankle_bp.xlsx") |> 
  mutate(diff = sbp_dp - sbp_tp)

descriptives(data = sbp, vars = diff, dens = TRUE)
```

While there is a large difference in blood pressure (around 60 mmHg) that warrants further checking, the curve is roughly symmetric with an approximately Normal distribution.

## Paired t-Test

To perform a paired t-test we will use the dataset `mod05_ankle_bp.xlsx`. We can perform a paired t-test using the `ttestPS()` function within the `jmv` package, where we defined the paired observations as: `pairs=list(list(i1 = 'variable1', i2 = 'variable2'))

```
ttestPS(data = sbp, pairs = list(list(i1 = "sbp_dp", i2 = "sbp_tp")),
        meanDiff = TRUE, ci = TRUE)
        

 PAIRED SAMPLES T-TEST

 Paired Samples T-Test                                                                                                                   
 ──────────────────────────────────────────────────────────────────────────
                                      statistic     df          p          
 ──────────────────────────────────────────────────────────────────────────
   sbp_dp    sbp_tp    Student's t    -0.9621117    106.0000    0.3381832  
 ──────────────────────────────────────────────────────────────────────────
 ────────────────────────────────────────────────────────────────────────────── 
                        Mean difference    SE difference    Lower        Upper      
 ────────────────────────────────────────────────────────────────────────────── 
   sbp_dp    sbp_tp        -1.261682         1.311368    -3.861596    1.338232   
 ────────────────────────────────────────────────────────────────────────────── 
   Note. Hₐ μ <sub>Measure 1 - Measure 2</sub> ≠ 0
```

The syntax of the ttestPS function is a little cumbersome. The `t.test` function can be used as an alternative:

```{r}
t.test(sbp$sbp_dp, sbp$sbp_tp, paired = TRUE)
```

